A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2022-12-22 paper/mcd2.tex
finalize proof of lemma on cd-add/mult

We will refer to \pr{cd-add} and \pr{cd-mult} as the \emph{basic CD iterations}.  Under appropriate assumptions on the damping, these iterations generate monotonically-decreasing objective function values whenever VI problem \eqref{eq:vi} corresponds to convex optimization as in problem \eqref{eq:minimization}.  In this sense the boxed VI problems in the basic CD iterations are analogous to constrained line searches; they reduce to line searches when the $\cV_i$ are one-dimensional.

@@

If $w$ is computed from \pr{cd-mult} then actually each iteration of the \textbf{for} loop lowers the objective value.  To show this let $z_{i} = \sum_{j\le i} w_j + \sum_{j > i} \Pi_j u \in \cK$ be the state at the end of the $i$th iteration.  Using convexity, observation \eqref{eq:admissibledifferencemult} with $v_i=\Pi_i u$, and the boxed inequality we find:
\begin{equation*}
F(z_{i-1}) - F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \ge \ip{f\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)}{\Pi_i u - \hat w_i} \ge 0.
\end{equation*}
Using convexity (again), and the above inequality, gives
\begin{align*}
F(z_i) &= F\left((1-\alpha) z_{i-1} + \alpha \left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)\right) \\
&\le  (1-\alpha) F(z_{i-1}) + \alpha F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \le F(z_{i-1}).
\end{align*}
Noting that $w = \sum_{i=0}^{m-1} w_i = z_{m-1}$, the result for \pr{cd-mult} follows.

FIXME $\alpha \le 1/m$ argument
\end{proof}

When $\alpha=1/m$ exactly then the logic of the proof of monotonicity for \pr{cd-add} is clearer.  In this case each VI solution over $\cK_i$ lowers the objective value from $F(u)$ to $F(u+e_i)$.  (As noted, these VI solutions can occur in parallel.)  Then the convex combination of points $\{u+e_i\} \subset \cK$, with equal coefficients $1/m$, is $w=u+\frac{1}{m}\sum_i e_i$.  By convexity of $F$ it follows that $F(w)$ is lower than the average $\frac{1}{m} \sum_i F(u+e_i)$, which is lower than $F(u)$.

>>>>

We will refer to \pr{cd-add} and \pr{cd-mult} as the \emph{basic CD iterations}.  According to damping assumptions, these iterations generate monotonically-decreasing objective values if VI problem \eqref{eq:vi} corresponds to convex optimization (problem \eqref{eq:minimization}), as shown in the following Lemma.  Thus the boxed VI problems in the \pr{cd-add} and \pr{cd-mult} algorithms are analogous to constrained line searches, and in fact they reduce to such when the $\cV_i$ are one-dimensional.

@@

Each iteration of the \textbf{for} loop in \pr{cd-mult} lowers the objective value.  To show this let $z_{i} = \sum_{j\le i} w_j + \sum_{j > i} \Pi_j u \in \cK$ be the state at the end of the $i$th iteration.  Using convexity, observation \eqref{eq:admissibledifferencemult} with $v_i=\Pi_i u$, and the boxed VI we find:
\begin{equation*}
F(z_{i-1}) - F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \ge \ip{f\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)}{\Pi_i u - \hat w_i} \ge 0.
\end{equation*}
Using convexity (again), and the above inequality, gives
\begin{align*}
F(z_i) &= F\left((1-\alpha) z_{i-1} + \alpha \left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)\right) \\
&\le  (1-\alpha) F(z_{i-1}) + \alpha F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \le F(z_{i-1}).
\end{align*}
Noting that $w = \sum_{i=0}^{m-1} w_i = z_{m-1}$, the result for \pr{cd-mult} follows.

Suppose $w=u+\alpha\sum_i e_i$ is computed by \pr{cd-add}.  Considering each $e_i=\hat w_i - \Pi_i u$ separately we have, by convexity and the boxed VI,
\begin{equation*}
F(u) - F(u+e_i) \ge \ip{f(u+e_i)}{u-(u+e_i)} = \ip{f(u-\Pi_i u + \hat w_i)}{\Pi_i u - \hat w_i} \ge 0.
\end{equation*}
If $z = u+ \frac{1}{m} \sum_i e_i$ then, again using convexity (Jensen's inequality),
\begin{align*}
F(u) - F(z) &= \frac{1}{m} \sum_{i=0}^{m-1} F(u) - F\left(u + \frac{1}{m} \sum_{i=0}^{m-1} e_i\right) \\
&\ge \frac{1}{m} \sum_{i=0}^{m-1} F(u+e_i) - F\left(\frac{1}{m} \sum_{i=0}^{m-1} (u+e_i)\right) \ge 0.
\end{align*}
Now suppose $\alpha \le 1/m$ and define $\lambda = m\alpha$ so $0<\lambda\le 1$ and $w = (1-\lambda) u + \lambda z$.  Thus by convexity and the above result, $F(w) \le (1-\lambda) F(u) + \lambda F(z) \le F(u)$.
\end{proof}

When $\alpha=1/m$ then the logic of the above proof for \pr{cd-add} is simpler.  Because each boxed VI over $\cK_i$ reduces the objective value from $F(u)$ to $F(u+e_i)$, we have $F(w)$ less than $\frac{1}{m} \sum_i F(u+e_i)$, which is lower than $F(u)$.
